Gravitational Lens Recovery with \Glass: Illustrating the power of free-form lens modelling

### 2013-06-12 limits/ms.tex
references added to intro (though more are still required); some polish added; theory outline sketched to be worked on this afternoon.

\documentclass[galley]{mn2e}
%\documentclass[twocolumn,galley]{mn2e}
%\documentclass[onecolumn,galley,draft]{mn2e}
%\usepackage{myaasmacros}
\usepackage{mathrsfs}
\usepackage{amsmath}
\usepackage{graphicx}

@@

%\title{Limits on constraining galaxy profiles with current and future strong gravitational lensing}
\title[\Glass]{Gravitational Lens Recovery with \Glass: Illustrating the power of free-form lens modelling}
\author{%
Jonathan P. Coles 
\and 
Justin I. Read
\and 
Prasenjit Saha 
}

@@

%\tableofcontents

\begin{abstract}
{\bf JR. Does \Glass\ stand for something?}\\
We introduce a new free-form gravitational lens modelling tool \Glass\ that builds on an earlier code \PixeLens. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects: (i) it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors; (ii) it is built on a fully modular framework that can support different basis functions; (iii) non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented); (iv) the stellar mass (if known) can be applied as a prior; and (v) adaptive resolution can be used in regions of interest. We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data are required to  measure the circularly averaged mass profile of a lens and its shape. We find that for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy. However, a single quad with time delays also performs well. Folding in additional constraints from stellar population synthesis modelling provides powerful additional constraints for galaxy lenses where the stars dominate the central potential, providing similarly good constraints from a single double without time delays as a quad with time delays in the absence of stellar mass constraints. Stellar kinematic data is only useful for probing the mass profile within the half light radius $r_{1/2}$ of the stars provided data are available for $r \gg r_{1/2}$ since this is required to integrate out the unknown velocity anisotropy $\beta(r)$. On the other hand, if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to probe cosmology. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
\end{abstract}

\section{Introduction}

Since the discovery of the first gravitational lens (CITE), lensing has become an increasingly important probe of the distribution of mass in the Universe (CITE). Strong lensing in particular provides a powerful probe of the mass distribution at the heart of galaxies and galaxy clusters given good enough data (CITE). This in turn provides valuable constraints on the nature of dark matter (CITE), on cosmological models (CITE), and on galaxy formation models (CITE). To date, some $\sim$400 strong lenses are known (CITE), but with the advent of large ground and space based surveys, this number is set to explode, with thousands of lenses expected to be discovered over the next ten years (CITE).

The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data (CITE). This is useful in that a simple model provides valuable physical insight (CITE). However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained (CITE). Instead of finding one `best-fit' model, we must search for many models consistent with the data, exploring degeneracies between different but plausible mass models. This latter word -- {\it plausible} -- encodes the key hidden gremlin in non-parametric methods: constraints extra to the available data are typically required on the solution space in order to make it finite -- priors (CITE). However, even rather restrictive priors are far less restrictive than the implicit constraints inherent in parametric mass modelling. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including addition non-lensing data) are required to measure parameters of interest.

In this paper, we present a new free-form or `non-parametric' strong lens modelling framework \Glass\ that builds on an earlier code \PixeLens (CITE). Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects:

\begin{enumerate} 
\item it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors;
\item it is built on a fully modular framework that can support different basis functions;
\item non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented);
\item the stellar mass (if known) can be applied as a prior; and
\item adaptive resolution can be used in regions of interest. 
\end{enumerate}

We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data from strong lensing, stellar populations synthesis models and/or stellar kinematics are required to measure the circularly averaged mass profile of a lens and its shape. We will consider using strong lensing as a cosmological probe, and present the first results from \Glass\ applied to real data, in forthcoming papers.

%(The SKA estimates that after its start in 2020 about 10,000 lenses will be found. The LSST expects to find about 5,000 multiply-imaged galaxies and about 150 AGNs or quasar lenses after 2022.)

%A typical lens, where only the measured observables are the image positions, can been used to estimate the total mass of a galaxy including parts of the dark matter halo. The enclosed mass inside the Einstein radius is simple to calculate and quite robust given the observed image positions. The exact mass distribution is more difficult to determine, however, as there are typically not enough constraints on the mass profile. The known degeneracies have been discussed at length in the literature. At the very least, one would like to have a measure of the enclosed mass at two distinct radii to be able to properly distinguish between analytic mass models. This would break the steepness degeneracy, also known as the mass sheet degeneracy.

%Lensing is not the only technique that can measure enclosed mass. The enclosed mass can also be estimated from stellar velocity dispersions. If the radius from this measurement is sufficiently different from the Einstein radius, then combining this measurement with lensing provides the necessary constraint. Another solution is to find a galaxy that is lensing two sources. The images from each source provide an enclosed mass. The extent to which the radii differ is dependent upon the redshifts of the two sources: the larger the redshift difference the larger the radial separation.

%Lensing is known to be sensitive to cosmology and, in principle, can be used to estimate such quantities as the Hubble constant $H_0$ when the lens has measurable time delays. However, due to strong degeneracies such as the steepness degeneracy there is often little one can learn from a single lens. In this case, one is free to change the mass and $H_0$ while keeping all observables fixed. A single lens can say almost nothing about the actual value of $H_0$. However, it has been shown that combining modeling results of several strong lenses with time delays can significantly reduce the allowed values and provides a measure to within 10 percent \cite{}.

%If one assumes a Hubble constant $H_0$ and time delays have been observed for a lens this will break the steepness degeneracy but will not help in constraining the mass profile at two radii.  One can choose to postpone this problem and simply fit models to the given data.  The earliest examples were singular isothermal sphere (SIS) or elliptical (SIE) models but have since evolved to multi-component models include separate models for the stars and dark matter. Other methods that do not ignore this problem use genetic algorithms or linear programming to explore the large space of solutions.

%Despite such difficulties lensing is a direct probe of the gravitating mass in a halo and there are a number of open problems in galaxy formation that lensing can help address. For instance, the debate remains as to what form the inner slope of the density profile takes. Numerical simulations of dark matter suggest a cusp, while observations from lensing and stellar dynamics suggest cores. This may, however, be only a matter of scale, as both theory and observations agree that clusters have cuspy profiles.

%Lensing may also be useful in measuring possible adiabatic contraction of the dark matter halo due to the central stars in a galaxy. For this, one needs high resolution mass models of the stellar profile combined with accurate models of the total mass. Indeed, with such models, one may also improve the statistics of the dark matter fraction in a galaxies over many different mass scales.

%These open problems suggest what kind of lensing data will be required in thefuture.  Future surveys will provide many lenses but time delays will require a long follow up. Stellar mass maps, however, take a known, fixed observation time and can be performed for many lenses. It will be important not only to measure the stellar velocity dispersion at an arbitrary radius, but to ensure that the measurement is taken at a radius different to the Einstein radius. For the radio surveys like SKA it will be important to detect the central image of the lens. This can provide mass information all the way to the center of the galaxy.

This paper is organised as follows. We first review the basic lensing theory we will need and then describe how \Glass\ is used to generate lensing models. The mock data and how it was generated is described in \secref{}. In section{} we discuss the lensing configurations in the parameter study and present some example output. Our finals results are presented in \secref{}.

%\subsection{Future} %-----------------------------------------------------------
    %\begin{itemize}
    %\item New satellites will find many new lenses.
    %\item LSST, DES, PanSTARRS?, SKA
    %\item What is the most important data to have?
    %\end{itemize}

%\subsection{Present} %----------------------------------------------------------
%    \begin{itemize}
%    \item SLACS, CASTLES, CosmoGrail, CLASS, CFHTLS, SNAP?
%    \end{itemize}

%\subsection{Why do we need accurate galaxy models?} %---------------------------
    %\begin{itemize}
    %\item Galaxy formation
    %\item Cusp/Core problem
    %\item Universal halos?
    %\item Adiabatic contraction
    %\item DM fraction
    %\end{itemize}

%\subsection{Past work on constraining profiles.} %------------------------------
    %\begin{itemize}
    %\item Much of the early literature used simple SIS and SIE parametric models. Understandable.
    %\item Bayesian model fitting
    %\item Free-form modeling
    %\item Keeton has a good review.
    %\end{itemize}

%\subsection{What are the problems?} %------------------------------
    %\begin{itemize}
    %\item Very little data per lens.
    %\item Several well-known degeneracies.
    %\item Velocity dispersions: aperture too small
    %\end{itemize}

%\subsection{Goals of the paper} %------------------------------
%    \begin{itemize}
%    \item Perform a parameter study over lens morphologies and quantify what they can tell us about profiles.
%    \end{itemize}

\section{Theoretical Background}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}

>>>>

\documentclass[galley]{mn2e}
%\documentclass[twocolumn,galley]{mn2e}
%\documentclass[onecolumn,galley,draft]{mn2e}
\usepackage{myaasmacros}
\usepackage{mathrsfs}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}

@@

\title[\Glass]{Gravitational Lens Recovery with \Glass: Illustrating the power of free-form lens modelling}
\author{%
Jonathan P. Coles 
\and 
Justin I. Read
\and 
Prasenjit Saha 
}

@@

\begin{abstract}
{\bf JR. Does \Glass\ stand for something?}\\
We introduce a new free-form gravitational lens modelling tool \Glass\ that builds on an earlier code \PixeLens. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects: (i) it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors; (ii) it is built on a fully modular framework that can support different basis functions; (iii) non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented); (iv) the stellar mass (if known) can be applied as a prior; and (v) adaptive resolution can be used in regions of interest. We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data -- from lensing, stellar kinematics and/or stellar masses -- are required to measure the circularly averaged mass profile of a lens and its shape. Our key findings are as follows: (i) for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy; (ii) a single quad with time delays also performs well, giving a good recovery of both the mass profile and its shape; (iii) stellar masses -- for lenses where the stars dominate the central potential -- can also break the steepness degeneracy, giving a recovery almost as good as having time delay data or multiple source redshifts; (iv) stellar kinematics are similarly powerful if the data extend beyond the half light radius of the stars $r \gg r_{1/2}$ (required to integrate out the effect of the unknown velocity anisotropy, $\beta(r)$); and (v) if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to constrain cosmological models. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
\end{abstract}

\section{Introduction}\label{sec:intro}
{\bf JR. Lots of references missing here. Need to be added. But this should get us started.}\\

Since the discovery of the first gravitational lens \citep{1979Natur.279..381W}, lensing has become an increasingly important probe of the distribution of mass in the Universe \citep{1937ApJ....86..217Z,2012arXiv1206.1225A}. Strong lensing in particular provides a powerful probe of the mass distribution at the heart of galaxies and galaxy clusters given good enough data \citep{2010CQGra..27w3001B,2007ApJ...667..645R,2006ApJ...652L...5S,2009ApJ...690..154S}. This in turn provides valuable constraints on the nature of dark matter \citep{2006ApJ...652L...5S,2013ApJ...765...25N}, on cosmological models \citep{2010Sci...329..924J, 2008ApJ...679...17C}, and on galaxy formation models \citep{2012MNRAS.424..104L}. To date, some $\sim$400 strong lenses are known\footnote{{\tt http://admin.masterlens.org/}.}, but with the advent of large ground and space based surveys, this number is set to explode, with thousands of lenses expected to be discovered over the next ten years \citep{2012arXiv1206.1225A, 2010AAS...21540115M, 2004NewAR..48.1085K}.

The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K, 2010GReGr..42.2151K}. This is useful in that a simple model provides valuable physical insight. However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Instead of finding one `best-fit' model, we must search for many models consistent with the data, exploring degeneracies between different but plausible\footnotemark\ mass models \citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including data from stellar population synthesis modelling, stellar kinematics, or other indepedent probes of the mass) are required to measure parameters of interest.

\footnotetext{The word {\it plausible} encodes the key hidden gremlin in non-parametric methods: constraints extra to the available data (priors) are typically required on the solution space in order to make it finite. However, even rather restrictive priors are far less restrictive than the implicit constraints inherent in parametric mass modelling.}

In this paper, we present a new free-form or `non-parametric' strong lens modelling framework \Glass\ that builds on an earlier code \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects:

\begin{enumerate} 
\item it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors \citep{2012MNRAS.425.3077L};
\item it is built on a fully modular framework that can support different basis functions;
\item non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented);
\item the stellar mass (if known) can be applied as a prior; and
\item adaptive resolution can be used in regions of interest. 
\end{enumerate}

We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data from strong lensing, stellar mass (derived from populations synthesis modelling) and/or stellar kinematics are required to measure the circularly averaged mass profile of a lens and its shape. We will consider using strong lensing as a cosmological probe, and present the first results from \Glass\ applied to real data, in forthcoming papers.

This paper is organised as follows. In \S\ref{sec:theory}, we review the basic lensing, stellar population synthesis, and Jeans mass modelling theory we will need. We present some simple back-of-the-envelope calculations that illustrate what type and quality of data we are likely to need to measure mass profiles and shapes. In \S\ref{sec:glass}, we describe the \Glass\ lens inversion algorithm and its stellar kinematic post-processing module. In \S\ref{sec:mockdata}, we describe our mock N-body data and our ray tracing algorithm used to generate the images. In \S\ref{sec:results}, we present our results from applying \Glass\ to these mock data. Finally, in \S\ref{sec:conclusions} we present our conclusions.

\section{Theoretical Background}\label{sec:theory}

\subsection{What can we learn from strong lensing data?}

\subsection{The utility of stellar mass estimates}

\subsection{Adding stellar kinematic constraints}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}\label{sec:glass}
